---
title: "Crizer 2023 LC PFAS Clint Data"
author: "John Wambaugh"
date: "April 15, 2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crizer 2024 Clint Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
wambaugh.john@epa.gov

# Here's the new R package for analyzing these data:
```{r load_packages, eval = TRUE}  
    library(invitroTKstats)
    
    # There are multiple packages for loading Excel files, but I've been using this
    # one lately:
    library(readxl)
```

    # Clear memory:
```{r clear_memory, eval = TRUE}     
    rm(list=ls())
```

Note that the following code chunks describe how the data.frame kreutz2023.clint
was created. Since we have already created this "level1" file we can just load
it from the invitroTKstats package. Any chunk that includes the statement 
"eval = FALSE" is not evaluated when the vignette is created.

    # Change to the directory that has your Excel file 
    # (this is where it is on my computer):
```{r set_working_directory, eval = TRUE}   
    setwd("c:/users/jwambaug/git/invitroTKstats/working/CrizerPFAS")
``` 
# Describe chemical tested:
```{r test_Chemicals, eval=TRUE}
ANALYSIS.LABEL <- "CrizerPFASApr2024"

chem.ids <- as.data.frame(read_excel("CrizerChemIDs-Apr2024.xlsx"))
```
# Load mass-spectrometry data:
```{r load_data, eval=TRUE}
data.guide <- as.data.frame(read_excel("dataguide-DC-hep-Apr2024.xlsx"))

dc.hep <- merge_level0(data.label=ANALYSIS.LABEL,
# describe the MS data table:
             level0.catalog=data.guide,
             type.colname.col="Type.ColName",
             istd.col="ISTD.Name",
             analysis.param.colname.col="RT.ColName",
             additional.colnames="Note",
             additional.colname.cols="Note.ColName",
# describe the chemical ID table:
             chem.ids=chem.ids,
             chem.lab.id.col="Chemical",
             chem.name.col="Chemical",
             num.rows.col="Num.Rows")
```

```{r annotate_data}
# Set reasonable sig figs on retention time:
dc.hep$Analysis.Params <- signif(as.numeric(dc.hep$Analysis.Params), 6)

# Get rid of NA notes:
dc.hep[is.na(dc.hep$Note),"Note"] <- ""

# Use invitroTKstats annotation of type for Crizer lab samples:
dc.hep <- subset(dc.hep,!is.na(Type))
dc.hep[regexpr("std",tolower(dc.hep[,"Type"]))!=-1, "Sample.Type"] <- "CC"
dc.hep[(regexpr("blank", tolower(dc.hep[,"Sample"])) != -1 &
       regexpr("media", tolower(dc.hep[,"Sample"])) != -1), "Sample.Type"] <- 
       "Blank"
dc.hep[regexpr("unknown",tolower(dc.hep[,"Type"]))!=-1, 
               "Sample.Type"] <- "Cvst"
dc.hep[regexpr("no_cells",tolower(dc.hep[,"Sample"]))!=-1, 
               "Sample.Type"] <- "Inactive"

# Use invitroTKstats annotation of type for Wetmore lab samples:
dc.hep[dc.hep$Type %in% c("Standard"), "Sample.Type"] <- "CC"
dc.hep[dc.hep$Type %in% c("Blank"), "Sample.Type"] <- "Blank"
for (this.string in c("t0-","t0 ","t15-","t15 ",
                      "t30-","t30 ","t60-","t60 ",
                      "t120-","t120 ","t240-", "t240 "))
  dc.hep[regexpr(this.string, tolower(dc.hep[,"Note"]))!=-1, 
               "Sample.Type"] <- "Cvst"
dc.hep[regexpr("inactive",tolower(dc.hep[,"Note"]))!=-1, 
               "Sample.Type"] <- "Inactive"
dc.hep[regexpr("wem",tolower(dc.hep[,"Note"]))!=-1, 
               "Sample.Type"] <- NA

dc.hep[regexpr("cc",tolower(dc.hep[,"Type"]))!=-1, "Sample.Type"] <- "CC"
dc.hep[regexpr("blank",tolower(dc.hep[,"Type"]))!=-1, "Sample.Type"] <- "Blank"
for (this.string in c("t0-","t0 ","t15-","t15 ",
                      "t30-","t30 ","t60-","t60 ",
                      "t120-","t120 ","t240-", "t240 "))
  dc.hep[regexpr(this.string, tolower(dc.hep[,"Type"]))!=-1, 
               "Sample.Type"] <- "Cvst"
dc.hep[regexpr("inactive",tolower(dc.hep[,"Type"]))!=-1, 
               "Sample.Type"] <- "Inactive"
dc.hep[regexpr("wem",tolower(dc.hep[,"Type"]))!=-1, 
               "Sample.Type"] <- NA

# Some of the Cvst data in this file were misannotated as blank:
dc.hep[dc.hep$Level0.File %in% 
         "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF.xlsx" &
       regexpr("Human Mix", dc.hep$Note) != -1, 
       "Sample.Type"] <- "Cvst" <- "Cvst"
# Some of the Cvst data in this file were misannotated as blank:
dc.hep[dc.hep$Level0.File %in% 
         "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF.xlsx" &
       regexpr("Inactive", dc.hep$Note) != -1, 
       "Sample.Type"] <- "Inactive"

# Figure out from sample name whether the hepatocytes were alive:
# Indicate whether hepatocytes have been inactivated:
dc.hep$Active.Hep <- NA

 
# Add time of sample:
dc.hep <- subset(dc.hep,!is.na(dc.hep[,"Sample"]))
dc.hep[,"Time"] <- NA
dc.hep[regexpr("no_cells",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 240/60
# Do this time first since zeros show up in other names:
dc.hep[regexpr("0min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 0/60
dc.hep[regexpr("240min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 240/60
dc.hep[regexpr("120min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 120/60
dc.hep[regexpr("90min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 90/60
dc.hep[regexpr("60min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 60/60
dc.hep[regexpr("30min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 30/60
dc.hep[regexpr("15min",tolower(dc.hep[,"Sample"]))!=-1,"Time"] <- 15/60

# Annotate time for Wetmore samples:
for (this.time in c(0,15,30,60,120,240))
  for (this.col in c("Note","Type"))
{
  this.time.str <- paste0("t",this.time,"-")
  dc.hep[regexpr(this.time.str,
                 tolower(dc.hep[,this.col]))!=-1,"Time"] <- this.time/60
  this.time.str <- paste0("t",this.time," ")
  dc.hep[regexpr(this.time.str,
                 tolower(dc.hep[,this.col]))!=-1,"Time"] <- this.time/60
}

# Differentiate live hepatocytes from inactive:
dc.hep[dc.hep[,"Sample.Type"] %in% "Cvst", "Active.Hep"] <- 1
dc.hep[dc.hep[,"Sample.Type"] %in% "Inactive", "Active.Hep"] <- 0
```

# Identify Wetmore lab chemicals:
```{r wetmore_chems}
wetmore.files <- c(
  "Hep13-Hep14-to-JFW_PFAS_REV.xlsx",
  "PFAS_20210902_Hep11_Data_to_JFW_20220318.xlsx",
  "20231106_HepCl_ReRun_KR.xlsx",
  "20220203_Hep12_SPEcassette_MGS-DataSummary.xlsx",
  "20240206_HepCl_KR_to_JFW-SDF.xlsx",
  "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF"
  )
```

# Set the dilution facor for the various samples:
```{r dilution_factors}
dc.hep$Dilution.Factor <- 1
dc.hep[!(dc.hep$Level0.File %in% wetmore.files) & 
       dc.hep[,"Sample.Type"] %in% c("Cvst","Inactive"), 
       "Dilution.Factor"] <- 4

dc.hep[dc.hep$Level0.File %in% wetmore.files,
  "Dilution.Factor"] <- 480

dc.hep[dc.hep$Level0.File %in% wetmore.files & 
       dc.hep$DTXSID %in% c(
  "DTXSID50226894",
  "DTXSID50379814",
  "DTXSID00880026"
  ),
  "Dilution.Factor"] <- 720

dc.hep[dc.hep$Level0.File %in% wetmore.files & 
       dc.hep$Sample.Type %in% c(
  "CC",
  "Blank"
  ),
  "Dilution.Factor"] <- 240

dc.hep[dc.hep$Level0.File %in% "20240206_HepCl_KR_to_JFW-SDF.xlsx" &
       dc.hep$DTXSID %in% c(
  "DTXSID3037709",
  "DTXSID40880025"
  ),
  "Dilution.Factor"] <- 36

dc.hep[dc.hep$Level0.File %in% "20240206_HepCl_KR_to_JFW-SDF.xlsx" &
       dc.hep$DTXSID %in% c(
  "DTXSID3038939"
  ),
  "Dilution.Factor"] <- 18
```


# Make sure the areas are numeric:
```{r numeric_peaks}
dc.hep$Peak.Area <- as.numeric(dc.hep$Peak.Area)
dc.hep$ISTD.Peak.Area <- as.numeric(dc.hep$ISTD.Peak.Area)
dc.hep[,"Compound.Conc"] <- as.numeric(dc.hep[,"Compound.Conc"])
```


# No ISTD for Crizer samples:
```{r no.istd}
dc.hep[!(dc.hep$Level0.File %in% wetmore.files),"ISTD.Peak.Area"] <- 1
```

# Second runs not used for Wetmore:
```{r not.really.blanks}
dc.hep[dc.hep$Type=="Blank" & 
       regexpr("blank",tolower(dc.hep$Note)) == -1 &
       regexpr("mix", tolower(dc.hep$Note)) == -1,
       "Sample.Type"] <- NA
```

# Convert nM standard concs to uM:
```{r nmtoum}
dc.hep[,"Compound.Conc"] <- as.numeric(dc.hep[,"Compound.Conc"])/1000
dc.hep[dc.hep$Sample.Type %in% c("Blank"), "Compound.Conc"] <- 0
```

```{r setinstruments}
dc.hep[!(dc.hep$Level0.File %in% wetmore.files),"Instrument"] <- "Waters Xevo TQ-S micro (QEB0036)"
dc.hep[(dc.hep$Level0.File %in% wetmore.files),"Instrument"] <- "Waters Xevo TQS-Micro UPLC-MS/MS"
```
# Create level 1 file:
```{r level1}
level1 <- format_clint(dc.hep,
  FILENAME=ANALYSIS.LABEL,
  sample.col ="Sample",
  date.col="Date",
  compound.col="Compound",
  lab.compound.col="Lab.Compound.ID",
  type.col="Sample.Type",
  dilution.col="Dilution.Factor",
  cal.col="Date",
  istd.conc = 10/1000,
  area.col="Peak.Area",
  istd.col= "ISTD.Peak.Area",
  density = 0.5,
  clint.assay.conc = 1,
  note.col="Note",
  std.conc.col="Compound.Conc",
  time.col = "Time",
  analysis.method = "UPLC-MS/MS",
  analysis.instrument.col = "Instrument",
  analysis.parameters.col = "Analysis.Params"
  )
```

```{r create_level2file}
level2 <- level1
level2[!(level2$Level0.File %in% wetmore.files), "Verified"] <- "Y"
level2[(level2$Level0.File %in% wetmore.files), "Verified"] <- "Y"

# Data file annotated wrt to verification using the note column:
level2[level2$Level0.Sheet %in% "Hep14-Raw-for JFW" &
       !(level2$Note %in% ""), "Verified"] <-
  level2[level2$Level0.Sheet %in% "Hep14-Raw-for JFW" &
       !(level2$Note %in% ""), "Note"]

# Bad T15 and T30 samples:
level2[level2$DTXSID %in% "DTXSID50381073" &
       level2$Lab.Sample.Name %in% c(
         "20220128_PFAS_Clint_Hep12WAX_Sample074",
         "20220128_PFAS_Clint_Hep12WAX_Sample075",
         "20220128_PFAS_Clint_Hep12WAX_Sample076",
         "20220128_PFAS_Clint_Hep12WAX_Sample081",
         "20220128_PFAS_Clint_Hep12WAX_Sample082",
         "20220128_PFAS_Clint_Hep12WAX_Sample083"
         ), "Verified"] <- "JFW: Early times inconsistent with later timepoints"



# Special annotation for DTXSID70565479
level2[level2$DTXSID %in% "DTXSID70565479" &
       level2$Lab.Sample.Name %in% c(
         "20220428_PFAS_Clint_Hep14_WAX_Sample005",
         "20220428_PFAS_Clint_Hep14_WAX_Sample006",
         "20220428_PFAS_Clint_Hep14_WAX_Sample012",
         "20220428_PFAS_Clint_Hep14_WAX_Sample013",
         "20220428_PFAS_Clint_Hep14_WAX_Sample014",
         "20220428_PFAS_Clint_Hep14_WAX_Sample016",
         "20220428_PFAS_Clint_Hep14_WAX_Sample017",
         "20220428_PFAS_Clint_Hep14_WAX_Sample018",
         "20220428_PFAS_Clint_Hep14_WAX_Sample020",
         "20220428_PFAS_Clint_Hep14_WAX_Sample021",
         "20220428_PFAS_Clint_Hep14_WAX_Sample022",
         "20220428_PFAS_Clint_Hep14_WAX_Sample024",
         "20220428_PFAS_Clint_Hep14_WAX_Sample025"
         ), "Verified"] <- "BAW: Ignore"

# 20220203_Hep12_SPEcassette_MGS-DataSummary.xlsx used two different mixtures of
# PFAS for testing, but the MS instrument ran in an automated mode, unaware that
# some chemicals were only in mixture "WAX1", and others were only in "WAX2":
level2[level2$Level0.File %in% 
         "20220203_Hep12_SPEcassette_MGS-DataSummary.xlsx" &
       level2$DTXSID %in% c(
         "DTXSID5061954",
         "DTXSID40880025",
         "DTXSID50381073") &
       regexpr("WAX1", level2$Note)!=-1, "Verified"] <- "Not in WAX1"
level2[level2$Level0.File %in% 
         "20220203_Hep12_SPEcassette_MGS-DataSummary.xlsx" &
       level2$DTXSID %in% c(
         "DTXSID80380256",
         "DTXSID90558000") &
       regexpr("WAX2", level2$Note)!=-1, "Verified"] <- "Not in WAX2"

# PFAS_20210902_Hep11_Data_to_JFW_20220318.xlsx used two different mixtures of
# PFAS for testing, but the MS instrument ran in an automated mode, unaware that
# some chemicals were only in mixture "WAX1", and others were only in "WAX2":
level2[level2$Level0.File %in% 
         "PFAS_20210902_Hep11_Data_to_JFW_20220318.xlsx" &
       level2$DTXSID %in% c(
         "DTXSID50226894",
         "DTXSID50379814",
         "DTXSID00880026") &
       regexpr("WAX1", level2$Note)!=-1, "Verified"] <- "Not in WAX1"
level2[level2$Level0.File %in% 
         "PFAS_20210902_Hep11_Data_to_JFW_20220318.xlsx" &
       level2$DTXSID %in% c(
         "DTXSID90382620",
         "DTXSID1032646") &
       regexpr("WAX2", level2$Note)!=-1, "Verified"] <- "Not in WAX2"


level2[level2$Level0.File %in% wetmore.files &
       regexpr("Mobile Phase Blank",level2$Note)!=-1, "Verified"] <- "Ignored"

level2[level2$Level0.File %in% wetmore.files &
       regexpr("WAX SPE Matrix Blank",level2$Note)!=-1, "Verified"] <- "Ignored"

level2[level2$Level0.File %in% wetmore.files &
       regexpr("WAX SPE Blank",level2$Note)!=-1, "Verified"] <- "Ignored"

level2[level2$Level0.File %in% wetmore.files &
       regexpr("WAX QC-A",level2$Note)!=-1, "Verified"] <- "Ignored"

level2[level2$Level0.File %in% wetmore.files &
       regexpr("WAX QC-B",level2$Note)!=-1, "Verified"] <- "Ignored"

level2[sapply(level2[,"Note"], function(x) "Excluded" %in% x), 
              "Verified"] <- "Excluded"
level2[sapply(level2[,"Note"], function(x) "Response Low" %in% x), 
              "Verified"] <- "Excluded"

# Three mixes used in "20231106_HepCl_ReRun_KR.xlsx":
# DTXSID70379917 and DTXSID8047553 were in Mix 1
these.chems <- c("DTXSID70379917", "DTXSID8047553")
this.mix <- "Mix 1"
level2[# Only want to touch data from this file:
       level2$Level0.File %in% "20231106_HepCl_ReRun_KR.xlsx" &
       # Only want to touch these chemicals:
       level2$DTXSID %in% these.chems &
       # Only want to touch measurements of mixtures:
       regexpr("Mix", level2$Note)!= -1 &
       # Get rid of measurements from wrong mixtures:
       regexpr(this.mix, level2$Note) == -1, "Verified"] <- "Only in Mix 1"
# DTXSID6067331 and DTXSID00192353 were in Mix 2
these.chems <- c("DTXSID6067331", "DTXSID00192353")
this.mix <- "Mix 2"
level2[# Only want to touch data from this file:
       level2$Level0.File %in% "20231106_HepCl_ReRun_KR.xlsx" &
       # Only want to touch these chemicals:
       level2$DTXSID %in% these.chems &
       # Only want to touch measurements of mixtures:
       regexpr("Mix", level2$Note)!= -1 &
       # Get rid of measurements from wrong mixtures:
       regexpr(this.mix, level2$Note) == -1, "Verified"] <- "Only in Mix 2"
# DTXSID50226894 and DTXSID5061954 were in Mix 3
these.chems <- c("DTXSID50226894", "DTXSID5061954")
this.mix <- "Mix 3"
level2[# Only want to touch data from this file:
       level2$Level0.File %in% "20231106_HepCl_ReRun_KR.xlsx" &
       # Only want to touch these chemicals:
       level2$DTXSID %in% these.chems &
       # Only want to touch measurements of mixtures:
       regexpr("Mix", level2$Note)!= -1 &
       # Get rid of measurements from wrong mixtures:
       regexpr(this.mix, level2$Note) == -1, "Verified"] <- "Only in Mix 3"

level2[level2$Lab.Sample.Name %in% c(
  "20220428_PFAS_Clint_Hep14_WAX_Sample005",
  "20220428_PFAS_Clint_Hep14_WAX_Sample006",
  "20220428_PFAS_Clint_Hep14_WAX_Sample012",
  "20220428_PFAS_Clint_Hep14_WAX_Sample013",
  "20220428_PFAS_Clint_Hep14_WAX_Sample014",
  "20220428_PFAS_Clint_Hep14_WAX_Sample016",
  "20220428_PFAS_Clint_Hep14_WAX_Sample017",
  "20220428_PFAS_Clint_Hep14_WAX_Sample018",
  "20220428_PFAS_Clint_Hep14_WAX_Sample020",
  "20220428_PFAS_Clint_Hep14_WAX_Sample021",
  "20220428_PFAS_Clint_Hep14_WAX_Sample022",
  "20220428_PFAS_Clint_Hep14_WAX_Sample024",
  "20220428_PFAS_Clint_Hep14_WAX_Sample025"), "Verified"] <- "Ignored"

# Get rid of rat data:
level2[level2$Level0.File %in% 
         "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF.xlsx" &
        regexpr("Rat", level2$Note) != -1, "Verified"] <- "Rat hepatocytes"

# Mixed chemicals:
this.mix <- "Mix 1/2"
chems.in.mix <- c("DTXSID30382063", "DTXSID00379268")
level2[level2$Level0.File %in% 
         "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF.xlsx" &
  regexpr(this.mix, level2$Note)!=-1 &
  !(level2$DTXSID %in% chems.in.mix),
    "Verified"] <- paste("Chemical not in", this.mix)
this.mix <- "Mix 3"
chems.in.mix <- c("DTXSID70191136")
level2[level2$Level0.File %in% 
         "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF.xlsx" &
  regexpr(this.mix, level2$Note)!=-1 &
  !(level2$DTXSID %in% chems.in.mix),
    "Verified"] <- paste("Chemical not in", this.mix)
this.mix <- "Mix 4/5"
chems.in.mix <- c("DTXSID50469320", "DTXSID60663110")
level2[level2$Level0.File %in% 
         "2023_09_19_Biotransformation_HepCl_KR_baw_to-JFW-SDF.xlsx" &
  regexpr(this.mix, level2$Note)!=-1 &
  !(level2$DTXSID %in% chems.in.mix),
    "Verified"] <- paste("Chemical not in", this.mix)

write.table(level2,
  file=paste0(ANALYSIS.LABEL,"-Clint-Level2.tsv"),
  sep="\t",
  row.names=F,
  quote=F)
```  

```{r plot_data}
for (this.id in unique(level2$DTXSID))
{
  this.subset <- subset(level2,DTXSID==this.id)
  try(plot(this.subset$Time,this.subset$Area,main=unique(this.subset$Compound.Name)))
}
```

```{r point_estimates}
level3 <- calc_clint_point(FILENAME=ANALYSIS.LABEL)
```

# repeat these bits in case a markov chain crashes and we need to restart:
```{r bayes_estimates}
library(invitroTKstats)
setwd("c:/users/jwambaug/git/invitroTKstats/working/CrizerPFAS")

level4 <- calc_clint(FILENAME=ANALYSIS.LABEL,
                          NUM.CORES=8,
                          decrease.prob = 0.05,
                          JAGS.PATH="C:/Users/jwambaug/AppData/Local/JAGS/JAGS-4.3.0/x64")  
``` 

```{r cvt_plots}
# Two different calibrations (order of magnitude difference in response) but same result except at final point. Final point for 042322 might make you think there is clearance, but is inconsistent with 111323.
plot_clint(level2,"DTXSID70379917")
# PFOA is an example of a chemical where the Bayes analysis thinks there’s degradation but no clearance. Looks to me like there may even have been a problem with the heat-treated controls. But I am comfortable with the no-clearance call here. The data look like this:
plot_clint(level2,"DTXSID8031865")
# A good no clearance call:
plot_clint(level2,"DTXSID8031863")
# Ammonium PFOA is an example where Bayes doesn’t think there is degradation, but also does not think there is clearance (but manual analysis found clearance). I’m torn on this one, since there are two phases – a decrease followed by a plateau. The Bayes analysis was torn, but I am guessing that since there were more points in the plateau that the Bayes model thought the early points were an anomaly:
plot_clint(level2,"DTXSID8037708")
# One where I am pretty confident in the Bayes is Perfluoro(2-ethoxyethane)sulfonic acid. The final points are very close between the heat-treated and active hepatocytes.
plot_clint(level2,"DTXSID50379814")
# Sodium perfluorodecanesulfonate seems like a good example where the Bayes method is just more conservative. You could argue that there is clearance, but the noise is such that I can buy it when the Bayes method says no clearance.
plot_clint(level2,"DTXSID60892443")
# The Bayes pValue for N-Ethylperfluorooctanesulfonamide is 0.0589. So it’s real close. But when I look at the data it looks like it degraded to me:
plot_clint(level2,"DTXSID1032646")
```